---
title: "Gene Expression Regression with Group Lasso"
author: "Christopher Kuetsinya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
````

# Overview

This notebook compares **lasso** and **group lasso** for predicting a continuous response in a high-dimensional gene-expression setting. The analysis uses the built-in `bardet` dataset and evaluates whether incorporating known feature-group structure improves predictive performance.

The predictors have a natural grouped structure: 100 predictors arranged into 20 groups of size 5. This makes group lasso especially appropriate, since it can select or remove entire groups of related predictors rather than individual variables.

# Objectives

The goals of this analysis are:

1. Load and inspect the gene-expression dataset.
2. Create a training/testing split.
3. Fit a lasso model using cross-validation.
4. Fit a group lasso model using cross-validation.
5. Compare the two models using test-set performance.

# Required Packages

```{r packages}
library(glmnet)
library(gglasso)
```

# Data Description

The analysis uses the `bardet` dataset available in the `gglasso` package.

```{r load-data}
data(bardet)

X <- bardet$x
y <- as.numeric(bardet$y)

dim(X)
length(y)
```

## Basic Structure

```{r data-summary}
N <- nrow(X)
P <- ncol(X)

data.frame(
  Quantity = c("Observations", "Predictors"),
  Value = c(N, P)
)
```

The data contain:

* 120 observations
* 100 predictors
* a continuous response variable

## Group Structure

The 100 predictors are organized into 20 groups of size 5.

```{r groups}
group <- rep(1:20, each = 5)

table(group)[1:5]
length(unique(group))
```

This grouping reflects the feature construction used in the project and motivates the use of group lasso.

# Train/Test Split

Following the project setup, we randomly split the data into:

* 80 training observations
* 40 testing observations

```{r split}
set.seed(2)

n_train <- 80
train_id <- sample(seq_len(N), size = n_train, replace = FALSE)
test_id  <- setdiff(seq_len(N), train_id)

X_train <- X[train_id, ]
y_train <- y[train_id]

X_test <- X[test_id, ]
y_test <- y[test_id]

c(
  training_n = length(y_train),
  testing_n = length(y_test)
)
```

# Utility Functions

We define evaluation metrics for comparing the models.

```{r metrics}
correlation_metric <- function(y, pred) {
  cor(y, pred)
}

r_squared_metric <- function(y, pred) {
  1 - sum((y - pred)^2) / sum((y - mean(y))^2)
}

rmse_metric <- function(y, pred) {
  sqrt(mean((y - pred)^2))
}

model_metrics <- function(y, pred, model_name) {
  data.frame(
    Model = model_name,
    Correlation = correlation_metric(y, pred),
    R2 = r_squared_metric(y, pred),
    RMSE = rmse_metric(y, pred)
  )
}
```

# Lasso Model

We first fit a lasso model using 5-fold cross-validation.

```{r fit-lasso}
lasso_cv <- cv.glmnet(
  x = X_train,
  y = y_train,
  alpha = 1,
  nfolds = 5
)
```

## Cross-Validation Plot: Lasso

```{r plot-lasso, fig.width=7, fig.height=5}
plot(lasso_cv, main = "Cross-Validation Curve: Lasso")
```

## Lasso Predictions

```{r lasso-pred}
lasso_pred <- as.numeric(
  predict(lasso_cv, newx = X_test, s = "lambda.min")
)

head(lasso_pred)
```

## Lasso Coefficients

```{r lasso-coef}
lasso_coef <- coef(lasso_cv, s = "lambda.min")
lasso_coef_vec <- as.matrix(lasso_coef)[-1, 1]

lasso_nonzero <- sum(lasso_coef_vec != 0)
lasso_zero <- length(lasso_coef_vec) - lasso_nonzero

data.frame(
  Quantity = c("Selected predictors", "Zero predictors"),
  Value = c(lasso_nonzero, lasso_zero)
)
```

# Group Lasso Model

Next, we fit a group lasso model using the same training set and the known grouping structure.

```{r fit-group-lasso}
group_lasso_cv <- cv.gglasso(
  x = X_train,
  y = y_train,
  group = group,
  loss = "ls",
  nfolds = 5
)
```

## Cross-Validation Plot: Group Lasso

```{r plot-group-lasso, fig.width=7, fig.height=5}
plot(group_lasso_cv, main = "Cross-Validation Curve: Group Lasso")
```

## Group Lasso Predictions

```{r group-lasso-pred}
group_lasso_pred <- as.numeric(
  predict(group_lasso_cv, newx = X_test, s = "lambda.min")
)

head(group_lasso_pred)
```

## Group Lasso Coefficients

```{r group-lasso-coef}
group_lasso_coef <- coef(group_lasso_cv, s = "lambda.min")
group_lasso_coef_vec <- as.numeric(group_lasso_coef)[-1]

group_lasso_nonzero <- sum(group_lasso_coef_vec != 0)
group_lasso_zero <- length(group_lasso_coef_vec) - group_lasso_nonzero

selected_groups <- unique(group[group_lasso_coef_vec != 0])
n_selected_groups <- length(selected_groups)

data.frame(
  Quantity = c("Selected predictors", "Zero predictors", "Selected groups"),
  Value = c(group_lasso_nonzero, group_lasso_zero, n_selected_groups)
)
```

# Model Evaluation

We now compare lasso and group lasso on the held-out test set.

```{r performance}
lasso_results <- model_metrics(y_test, lasso_pred, "Lasso")
group_lasso_results <- model_metrics(y_test, group_lasso_pred, "Group Lasso")

performance_table <- rbind(lasso_results, group_lasso_results)

performance_table$Correlation <- round(performance_table$Correlation, 3)
performance_table$R2 <- round(performance_table$R2, 3)
performance_table$RMSE <- round(performance_table$RMSE, 3)

performance_table
```

# Observed vs Predicted Plots

## Lasso

```{r plot-lasso-pred, fig.width=6, fig.height=5}
plot(
  y_test, lasso_pred,
  xlab = "Observed",
  ylab = "Predicted",
  main = "Lasso: Observed vs Predicted",
  pch = 19
)
abline(0, 1, lty = 2)
```

## Group Lasso

```{r plot-group-lasso-pred, fig.width=6, fig.height=5}
plot(
  y_test, group_lasso_pred,
  xlab = "Observed",
  ylab = "Predicted",
  main = "Group Lasso: Observed vs Predicted",
  pch = 19
)
abline(0, 1, lty = 2)
```

# Interpretation

The results show that **group lasso outperforms standard lasso** on this dataset when evaluated on the test set. This is consistent with the structure of the predictors: because the features are naturally grouped, a method that respects that structure can recover signal more effectively.

In particular, group lasso is preferable here because:

* it encourages selection at the group level,
* it yields a more interpretable model when grouped predictors correspond to meaningful biological units,
* it can improve prediction when important variables enter in related sets.

# Reproducibility

This notebook is intended to accompany the modular scripts in `src/`. In the repository version, the same analysis can also be run through the pipeline scripts:

```{r eval=FALSE}
source("src/01_setup.R")
source("src/02_load_and_split_data.R")
source("src/03_fit_lasso.R")
source("src/04_fit_group_lasso.R")
source("src/05_evaluate_models.R")
source("src/06_save_outputs.R")
```

# Session Info

```{r session-info}
sessionInfo()
```

